# Regressions using survey data

## Literature summary

- Endogenous sampling can be thought of as cases where the regression error term is related to the sampling criteria [@friedman_tools_2013, @solon_what_2013, @fuller_sampling_2009]
- In the presence of endogenous sampling, unweighted estimates may be biased, but will be corrected by weighting by the inverse probability of selection [@friedman_tools_2013, @solon_what_2013, @fuller_sampling_2009]
- In the presence of endogenous sampling, if the sampling probability varies across certain strata and those strata indicators are included in the estimating equation, then the probability of selection should no longer be related to the error term. Subsequently, weighting is not necessary [@friedman_tools_2013, @solon_what_2013, @fuller_sampling_2009]
- In the case of a linear regression model that correctly specifies the conditional mean, the sampling would be exogenous if the sampling probabilities are independent of the error term in the regression equation. This would be the case, for example, if the sampling probabilities vary only on the basis of explanatory variables [@solon_what_2013]
- More generally, the issue is whether the sampling is independent of the dependent variable conditional on the explanatory variables [@solon_what_2013] 
- Weighting does not correct for confounding. Adjusting for confounders is still necessary.
- Weighting is useful when the regression model that accounts for sampling probabilities does not make any sense (e.g. in the case-control scenario).

## Setup

```{r setup}
FormatNicely <- function(x,dp=3){
  formatC(x,digits=dp,format="f")
}

ConvertFitToResults <- function(fit,dataName,analysisName){
  r <- data.frame(coef(summary(fit)))[-1,-c(3:4),drop=F]
  r[,1] <- sprintf("%s [%s]",
                   FormatNicely(r[,1],dp=3),
                   FormatNicely(r[,2],dp=3))
  r <- r[,1,drop=F]
  r <- as.data.frame(t(r))
  r$data <- dataName
  r$analysis <- analysisName
  return(r)
}
```


## Case-control studies

In this case study we will create a dataset `popData` that has a 1:1 linear relationship between the continuous exposure `x` and and the continuous outcome `y`. We also dichotomise `y` into a binary outcome `yBinary`.

We then create a dataset `casecontrolData` that oversamples cases 5x higher than is found in the normal population `popData`. We will then run linear regressions in the original population dataset and the case-control dataset and see how the effect estimates are affected by oversampling cases.

```{r}
library(data.table)

# Creating population dataset
x <- runif(100000)
popData <- data.table(x)
popData[,y:=x+rnorm(.N)]

# Binary outcome
popData[,yBinary:=0]
popData[y>1,yBinary:=1]
popData[,oversampled:=yBinary]

popData[,inclusionProb := 5]
popData[oversampled==0, inclusionProb:=1]

# Case control dataset
cases <- popData[yBinary==1]
controls <- popData[yBinary==0]

# Cases are sampled 5x higher than normal
casecontrolData <- rbind(cases,cases,cases,cases,cases,controls)
casecontrolData[,id:=1:.N]
```

```{r}
# Full population data set, unweighted regression
res <- vector("list",10)
fit <- glm(y~x,data=popData)
res[[1]] <- ConvertFitToResults(fit,dataName="Pop",analysisName="Unweighted")
res[[1]]$correlation <- FormatNicely(cor(resid(fit),fit$data$oversampled), dp=2)

# Biased dataset (case control with 5x), unweighted regression
fit <- glm(y~x,data=casecontrolData)
res[[2]] <- ConvertFitToResults(fit,dataName="Sample",analysisName="Unweighted")
res[[2]]$correlation <- FormatNicely(cor(resid(fit),fit$data$oversampled), dp=2)

# Biased dataset (case control with 5x), weighted regression
des <- survey::svydesign(id=~id,prob=~inclusionProb,data=casecontrolData)
fit <- survey::svyglm(y~x, design=des)
res[[3]] <- ConvertFitToResults(fit,dataName="Sample",analysisName="Weighted")
res[[3]]$correlation <- FormatNicely(cor(resid(fit),fit$data$oversampled), dp=2)

res <- rbindlist(res,fill=T)
setcolorder(res, c("data", "analysis", "x", "correlation"))
setnames(res, c("Data", "Analysis", "coef(x) [sd(coef(x))]", "Correlation*"))

knitr::kable(
  res, booktabs = TRUE,
  caption = 'Effects of weights on linear regression coefficient estimates. (*Correlation between residuals and sampling probability).'
)
```

We can see here that in the presence of endogenous sampling, the sampling probability is highly correlated with the regression error term. By weighting the data by the inverse probability of selection we obtain unbiased estimates of `coef(x)`.

## Oversampling a population with a higher level of the outcome

```{r}
library(data.table)

# Creating population dataset
x <- runif(100000)
popData <- data.table(x)
popData[,poor:=0]
popData[1:10000,poor:=1]
popData[,bmi:=22+1*x+5*poor+rnorm(.N)*2]

# Oversampled poor dataset
poor <- popData[poor==1]
notpoor <- popData[poor==0]

# Poor people are sampled 5x higher than not-poor
oversampledData <- rbind(poor,poor,poor,poor,poor,notpoor)
oversampledData[,id:=1:.N]

# Probability of inclusion
oversampledData[,inclusionProb := 5]
oversampledData[poor==0, inclusionProb:=1]
```

```{r}
# Full population data set, unweighted regression
res <- vector("list",10)
res[[1]] <- ConvertFitToResults(fit <- glm(bmi~x,data=popData),
                                dataName="Pop",analysisName="Unweighted")
res[[1]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), unweighted regression
res[[2]] <- ConvertFitToResults(fit <- glm(bmi~x,data=oversampledData),
                                dataName="Sample",analysisName="Unweighted")
res[[2]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), weighted regression
des <- survey::svydesign(id=~id,prob=~inclusionProb,data=oversampledData)
res[[3]] <- ConvertFitToResults(fit <- survey::svyglm(bmi~x, design=des),
                                dataName="Sample",analysisName="Weighted")
res[[3]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Full population data set, unweighted regression + strata indicator
res[[4]] <- ConvertFitToResults(fit <- glm(bmi~x+poor,data=popData),
                                dataName="Pop",analysisName="Unweighted+Strata")
res[[4]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), unweighted regression + strata indicator
res[[5]] <- ConvertFitToResults(fit <- glm(bmi~x+poor,data=oversampledData),
                                dataName="Sample",analysisName="Unweighted+Strata")
res[[5]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), weighted regression + strata indicator
des <- survey::svydesign(id=~id,prob=~inclusionProb,data=oversampledData)
res[[6]] <- ConvertFitToResults(fit <- survey::svyglm(bmi~x+poor, design=des),
                                dataName="Sample",analysisName="Weighted+Strata")
res[[6]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

res <- rbindlist(res,fill=T)
setcolorder(res,c("data", "analysis", "x", "poor", "correlation"))
setnames(res,c("Data", "Analysis", "coef(x) [sd(coef(x))]", "coef(poor) [sd(coef(poor))]", "Correlation*"))

knitr::kable(
  res, booktabs = TRUE,
  caption = 'Effects of weights on linear regression coefficient estimates. (*Correlation between residuals and sampling probability).'
)
```

We can see here that in the first three models the sampling probability is highly correlated with the regression error term. Models 1 and 3 provide unbiased estimates (through exogenous sampling and weighting, respectively.) In models 4 through to 6, we included an explanatory variable `poor` to account for the varying sampling probabilities. Subsequently, the sampling probability is no longer correlated with the regression error term, and we obtain unbiased estimates.

## Oversampling a population with a higher level of the outcome and exposure

```{r}
library(data.table)

# Creating population dataset
x <- runif(100000)
popData <- data.table(x)
popData[,poor:=0]
popData[1:10000,poor:=1]
popData[,x:=x+2*poor]
popData[,bmi:=22+1*x+5*poor+rnorm(.N)*2]

# Oversampled poor dataset
poor <- popData[poor==1]
notpoor <- popData[poor==0]

# Poor people are sampled 5x higher than not-poor
oversampledData <- rbind(poor,poor,poor,poor,poor,notpoor)
oversampledData[,id:=1:.N]

# Probability of inclusion
oversampledData[,inclusionProb := 5]
oversampledData[poor==0, inclusionProb:=1]
```

```{r}
# Full population data set, unweighted regression
res <- vector("list",10)
res[[1]] <- ConvertFitToResults(fit <- glm(bmi~x,data=popData),
                                dataName="Pop",analysisName="Unweighted")
res[[1]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), unweighted regression
res[[2]] <- ConvertFitToResults(fit <- glm(bmi~x,data=oversampledData),
                                dataName="Sample",analysisName="Unweighted")
res[[2]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), weighted regression
des <- survey::svydesign(id=~id,prob=~inclusionProb,data=oversampledData)
res[[3]] <- ConvertFitToResults(fit <- survey::svyglm(bmi~x, design=des),
                                dataName="Sample",analysisName="Weighted")
res[[3]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Full population data set, unweighted regression + strata indicator
res[[4]] <- ConvertFitToResults(fit <- glm(bmi~x+poor,data=popData),
                                dataName="Pop",analysisName="Unweighted+Strata")
res[[4]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), unweighted regression + strata indicator
res[[5]] <- ConvertFitToResults(fit <- glm(bmi~x+poor,data=oversampledData),
                                dataName="Sample",analysisName="Unweighted+Strata")
res[[5]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

# Biased dataset (poor oversampled 5x), weighted regression + strata indicator
des <- survey::svydesign(id=~id,prob=~inclusionProb,data=oversampledData)
res[[6]] <- ConvertFitToResults(fit <- survey::svyglm(bmi~x+poor, design=des),
                                dataName="Sample",analysisName="Weighted+Strata")
res[[6]]$correlation <- FormatNicely(cor(resid(fit),fit$data$poor), dp=2)

res <- rbindlist(res,fill=T)
setcolorder(res,c("data", "analysis", "x", "poor", "correlation"))
setnames(res,c("Data", "Analysis", "coef(x) [sd(coef(x))]", "coef(poor) [sd(coef(poor))]", "Correlation*"))

knitr::kable(
  res, booktabs = TRUE,
  caption = 'Effects of weights on linear regression coefficient estimates. (*Correlation between residuals and sampling probability).'
)
```

We can see here that in the first three models the sampling probability is highly correlated with the regression error term. Models 1 and 3 provide estimates unbiased due to the endogenous sampling (through exogenous sampling and weighting, respectively), however, these estimates are still biased due to confounding. In models 4 through to 6, we included an explanatory variable `poor` to account for the varying sampling probabilities. Subsequently, the sampling probability is no longer correlated with the regression error term, and we obtain unbiased estimates.
